Fast Regression of the Tritium Breeding Ratio in Tokamak Fusion Reactors

### 2020-06-11 paper/conclusion.tex
implement first batch of comments from Nikos

We employed a broad spectrum of data
analysis and machine learning techniques to develop fast and high-quality
surrogates for the Paramak TBR model. After reviewing
nine surrogate model families, examining their behaviour on constrained and
unrestricted feature space, and studying their scaling properties, we retrained
the best-performing instances to produce properties desirable for
practical use. The fastest surrogate, an artificial neural network trained
on~\num{500000} datapoints, featured~$R^2=\num{0.985}$ with mean prediction time
of~$\SI{0.898}{\micro\second}$, representing a relative
speedup of $8\cdot 10^6$ with respect to the MC model. We demonstrated the possibility of achieving comparable results using only a
training set of size~\num{10000}.

We further developed a novel adaptive
sampling algorithm, QASS, capable of interfacing with any of the studied
surrogates. Preliminary testing on a toy theory, qualitatively comparable to
the MC TBR model, demonstrated the effectiveness of QASS and behavioural trends
consistent with the design of the algorithm. With~\num{100000} initial samples and 100 incremental samples per iteration, QASS achieved a ${\sim}40\%$ decrease in surrogate error as compared with a baseline random sampling scheme. Further optimisation over the hyperparameter space has strong potential to increase this performance, in particular by decreasing the required quantity of initial samples. This will allow for future deployment of QASS on the Paramak TBR model in coalition with any of the most effective identified surrogates.

>>>>

We employed a broad spectrum of data
analysis and machine learning techniques to develop fast and high-quality
surrogates for the Paramak TBR model. After reviewing
nine surrogate model families, examining their behaviour on a constrained and
unrestricted feature space, and studying their scaling properties, we retrained
the best-performing instances to produce properties desirable for
practical use. The fastest surrogate, an artificial neural network trained
on~\num{500000} datapoints, attained an~$R^2=\num{0.985}$ with mean prediction time
of~$\SI{0.898}{\micro\second}$, representing a relative
speedup of $8\cdot 10^6$ with respect to the MC model. We demonstrated the possibility of achieving comparable results using only a
training set of size~\num{10000}.

We further developed a novel adaptive
sampling algorithm, QASS, capable of interfacing with any surrogate model.
Preliminary testing on a toy theory, qualitatively comparable to
the MC TBR model, demonstrated the effectiveness of QASS and behavioural trends
consistent with the design of the algorithm. With~\num{100000} initial samples
and 100 incremental samples per iteration, QASS achieved a ${\sim}40\%$ decrease
in surrogate error compared to a baseline random sampling scheme. Further optimisation over the hyperparameter space has strong potential to increase this performance, in particular by decreasing the required quantity of initial samples. This will allow for future deployment of QASS on the Paramak TBR model in coalition with any of the most effective identified surrogates.

### 2020-06-11 paper/methodology.tex
implement first batch of comments from Nikos

To compare quality of produced surrogates, we define a number of metrics listed
in~\Tref{tbl:metrics}. For regression performance analysis, we include a
selection of absolute metrics to assess their approximation capability and set
practical bounds on the expected uncertainty of their predictions. In addition, we also track
relative measures that are better-suited for comparison between this work and others as
they maintain invariance with respect to the selected domain and image space.
For complexity analysis, surrogates are assessed in terms of wall
time (captured by the Python~\texttt{time} package). This is motivated by common practical use
cases of our work, where models are trained and used as drop-in replacements for the
expensive MC TBR model. Since training set sizes remain to be determined, all times are
reported per a single datapoint. Even though some surrogates support acceleration
by means of parallelisation, sequential processing of samples was ensured to
achieve comparability between considered models. The only exception to this are
ANNs, which require considerable amount of processing power for training on
conventional CPU architectures. Lastly, to prevent undesirable bias by training set
selection, all metrics are collected in the scheme of 5-fold cross-validation.

@@

Experiments related to the decoupled approach are organised in four parts,
further described in this section. In summary, we aim to optimise hyperparameters of
each surrogate family separately and later use the best found results to compare surrogate
families among themselves.

The objective of the first experiment is to simplify the regression task for
surrogates prone to suboptimal performance in discontinuous spaces.
To this end, training points are filtered to a single selected discrete feature
assignment, and surrogates are trained only on the remaining continuous features.
This is repeated for 4~distinct assignments to explore variances in behaviour.
The second experiment conventionally measures surrogate performance on the full,
unrestricted feature space. In both cases, hyperparameter tuning is
facilitated by Bayesian optimisation~\cite{movckus1975bayesian}. We set its
objective to maximise $R^2$ and terminate the process after 1000~iterations or
2~days, whichever condition is satisfied first.

>>>>

To compare the quality of the produced surrogates, we define a number of metrics listed
in~\Tref{tbl:metrics}. For regression performance analysis, we include a
selection of absolute metrics to assess the models' approximation capability and set
practical bounds on the expected uncertainty of their predictions. In addition, we also track
relative measures that are better-suited for comparison between this work and others as
they are invariant with respect to the selected domain and image space.
For complexity analysis, surrogates are assessed in terms of wall
time (captured by the Python~\texttt{time} package). This is motivated by common practical use
cases of our work, where models are trained and used as drop-in replacements for the
expensive MC TBR model. Since training set sizes remain to be determined, all times are
reported per a single datapoint. Even though some surrogates support acceleration
by means of parallelisation, sequential processing of samples was ensured to
achieve comparability between considered models. The only exception to this are
ANNs, which require considerable amount of processing power for training on
conventional CPU architectures. Lastly, to prevent undesirable bias by training set
selection, all metrics are collected in the scheme of 5-fold cross-validation.

@@

Experiments related to the decoupled approach are organised in four parts,
further described in this section. In summary, we aim to optimise the hyperparameters of
each surrogate family separately and later use the best found results to compare surrogate
families among themselves.

The objective of the first experiment is to simplify the regression task for
surrogates prone to suboptimal performance in discontinuous spaces.
To this end, training points are filtered to a single selected discrete feature
assignment, and surrogates are trained only on the remaining continuous features.
This is repeated for 4~distinct assignments to explore variances in behaviour.
The second experiment conventionally measures surrogate performance on the full,
unrestricted feature space. In both cases, hyperparameter tuning is
facilitated by Bayesian optimisation~\cite{movckus1975bayesian}. We select the
hyperparameter configuration that produces the model that maximises $R^2$. The
process is terminated after 1000~iterations or
2~days, whichever condition is satisfied first.

### 2020-06-11 paper/results.tex
implement first batch of comments from Nikos

The results displayed in~\Fref{fig:exp1-time-vs-reg} indicate that in the first,
simplified case GBTs clearly appear to be the most accurate as
well as the fastest surrogate family in terms of mean prediction time. Following
that, we note that ERTs, SVMs and ANNs also achieved satisfactory results with respect to both examined metrics.
While the remainder of tested surrogate families does not exhibit problems in
complexity, its regression performance falls below average.

@@

Comparing these results with those of the second, unrestricted experiment (shown
in~\Fref{fig:exp2-time-vs-reg}), we observe that many surrogate families
consistently underperformed. The least
affected models appear to be GBTs, ANNs and ERTs, which are known to be capable of capturing relationships
involving mixed feature types that were deliberately withheld in the first
experiment. With only negligible differences, the first two of these families
appear to be tied for the best performance as well as the shortest prediction
time. We observe that ERTs and RBFs also
demonstrated satisfactory results, relatively outperforming the remaining surrogates in
terms of regression performance, and in some cases also in prediction time.

@@

Following both hyperparameter tuning experiments, we conclude that while domain
restrictions employed in the first case have proven effective in improving the
regression performance of some methods, this result has fluctuated considerably
depending on the selected slices. Furthermore, in all instances the best
results were achieved by families of surrogates that were nearly unaffected by
this modification.

@@

The results shown in~\Fref{fig:scaling} suggest that in terms of regression
performance the most accurate families from the previous experiments
consistently maintain their relative advantage over others, even as
more training points are introduced. While such families achieve nearly comparable
performance on the largest dataset, in the opposite case tree-based approaches
clearly outperform ANNs. This can be observed
particularly on sets of sizes up to~\num{6000}.

@@

Consistent with our expectation, the shortest training times were achieved by
instance-based learning methods (KNN, IDW) that
are trained trivially at the expense of increased lookup complexity later during prediction.
Furthermore, we observe that the majority of tree-based algorithms also perform
and scale well, unlike RBFs and GPR which appear to behave superlinearly. We note that ANNs,
which are the only family to utilise parallelisation during training, show an
inverse scaling characteristic. We suspect that this effect may be caused
by a constant multi-threading overhead that dominates the training process
on relatively small sets.

Finally, all tested families with the exception of previously mentioned instance-based
models offer desirable prediction times. Analogous to previous experiments,
GBTs, ABTs and ANNs appear to be tied, as they not only exhibit
comparable times but also similar scaling slopes. Following that, we note a
clear hierarchy of ERTs, SVMs, GPR and RBFs, trailed by IDW and KNNs.

\subsubsection{Model Comparison}
In the fourth experiment, we aim to create models that yield:
(a)~the best regression performance regardless
of other features, (b)~acceptable performance with the shortest mean
prediction time, or (c)~acceptable performance with the smallest training set.
To this end, we trained 8~surrogates that are presented in~\Fref{fig:reg-performance}
and~\Tref{tbl:exp4-detailed-results}.
Having selected ANNs, GBTs, ERTs, RBFs and SVMs based on the results of
experiments~2-3, we utilised the best-performing hyperparameters.
In pursuit of goal~(a), the best approximator (no.~1,
ANN) achieved~$R^2=\num{0.998}$ and mean prediction
time~$\overline{t}_{\text{pred.}}=\SI{1.124}{\micro\second}$. These correspond
to a standard error~$S=\num{0.013}$ and a relative speedup~$\omega=\num{6916416} \times$
with respect to the MC TBR evaluation baseline. Satisfying
goal~(b), the fastest model (no.~2, ANN) achieved~$R^2=\num{0.985}$,
$\overline{t}_{\text{pred.}}=\SI{0.898}{\micro\second}$, $S=\num{0.033}$
and~$\omega=\num{8659251} \times$.
While these surrogates
were trained on the entire available set of~\num{500000} datapoints, to satisfy
goal~(c) we also trained a more simplified model (no.~4, GBT)
that achieved~$R^2=\num{0.913}$,
$\overline{t}_{\text{pred.}}=\SI{6.125}{\micro\second}$, $S=\num{0.072}$ and $\omega=\num{1269777} \times$
with a set of size only~\num{10000}.

\begin{table*}
	\centering
	\sisetup{round-mode=places,round-precision=3,detect-weight=true,detect-family=true}
	\caption{\label{tbl:exp4-detailed-results}Results of experiment~4. Here, figures are reported over 5~cross-validation folds,
		$|\mathcal{T}|$~denotes cross-validation set size ($\times 10^3$)
		and $\omega$ is a relative speedup with respect to
		$\overline{t}_{\text{eval.}}=\SI{7.777049573054314}{\second}$
		measured in the MC TBR model during run~1 (see~\Tref{tbl:sampling-runs}).
		The best-performing metrics are highlighted in bold.}
	\setlength\tabcolsep{4pt}
	\begin{indented}
	\item[]
		\scriptsize
		\begin{tabular}{lrrrrrrrr}
		\toprule
		{} & {} & \multicolumn{4}{c}{Regression performance} &
		\multicolumn{3}{c}{Complexity}\\
		\cmidrule(lr){3-6}
		\cmidrule(lr){7-9}
		Model & $|\mathcal{T}|$ & MAE [TBR] & $S$ [TBR] & $R^2$ [rel.] & $R^2_{\text{adj.}}$ [rel.]
						& $\overline{t}_{\text{trn.}}$ [\si{\milli\second}] &
		$\overline{t}_{\text{pred.}}$ [\si{\milli\second}] & $\omega$ [rel.]\\
		\midrule
		\input{tbl/table11}
		\bottomrule
		\end{tabular}
	\end{indented}
\end{table*}

@@

	\caption{Regression performance of models 1-8 (from left to right, top to
		bottom) trained in the model comparison experiment, viewed
		as true vs.~predicted TBR on a test set of a selected cross-validation
		fold. Points are coloured by density.}
	\label{fig:reg-performance}
\end{figure}

>>>>

The results displayed in~\Fref{fig:exp1-time-vs-reg} indicate that in the first,
simplified case GBTs clearly appear to be the most accurate as
well as the fastest surrogate family in terms of mean prediction time. Following
that, we note that ERTs, SVMs and ANNs also achieved satisfactory results with respect to both examined metrics.
While the remainder of tested surrogate families does not exhibit prohibitive
complexity, its regression performance falls below average.

@@

Comparing these results with those of the second, unrestricted experiment (shown
in~\Fref{fig:exp2-time-vs-reg}), we observe that many surrogate families
consistently underperform. The least
affected models appear to be GBTs, ANNs and ERTs, which are known to be capable of capturing relationships
involving mixed feature types that were deliberately withheld in the first
experiment. With only negligible differences, the first two of these families
appear to be tied for the best performance as well as the shortest prediction
time. We observe that ERTs and RBFs also
demonstrated satisfactory results, clearly outperforming the remaining surrogates in
terms of regression performance, and in some cases also in prediction time.

@@

Following both hyperparameter tuning experiments, we conclude that while domain
restrictions employed in the first case have proven effective in improving the
regression performance of some methods, their performance fluctuates considerably
depending on the selected slices. Furthermore, in all instances the best
results are achieved by families of surrogates that do not benefit from this
restriction (GBTs, ANNs, ERTs).

@@

The results shown in~\Fref{fig:scaling} suggest that in terms of regression
performance the most accurate families from the previous experiments
consistently maintain their relative advantage over others, even as
more training points are introduced. While such families achieve nearly comparable
performance on the largest dataset, in the opposite case tree-based ensemble approaches
clearly outperform ANNs. This can be observed
particularly on sets of sizes up to~\num{6000}.

@@

Consistent with our expectation, the shortest training times were achieved by
instance-based learning methods (KNN, IDW) that
are trained trivially at the expense of increased lookup complexity later during prediction.
Furthermore, we observe that the majority of tree-based ensemble algorithms also perform
and scale well, unlike RBFs and GPR which appear to behave superlinearly. We note that ANNs,
which are the only family to utilise parallelisation during training, show an
inverse scaling characteristic. We suspect that this effect may be caused
by a constant multi-threading overhead that dominates the training process
on relatively small sets.

Finally, all tested families with the exception of previously mentioned instance-based
models offer desirable prediction times. Analogous to previous experiments,
GBTs, ABTs and ANNs appear to be tied, as they not only exhibit
comparable times but also similar scaling slopes. Following those, we note a
clear hierarchy of ERTs, SVMs, GPR and RBFs, trailed by IDW and KNNs.

\subsubsection{Model Comparison}
In the fourth experiment, we aim to create models that yield:
(a)~the best regression performance regardless
of other features, (b)~acceptable performance with the shortest mean
prediction time, or (c)~acceptable performance with the smallest training set.
To this end, we trained 8~surrogates that are presented in~\Fref{fig:reg-performance}
and~\Tref{tbl:exp4-detailed-results}.
Having selected ANNs, GBTs, ERTs, RBFs and SVMs based on the results of
experiments~2-3, we utilised the best-performing hyperparameters.
In pursuit of goal~(a), the best approximator (Model~1,
ANN) achieved~$R^2=\num{0.998}$ and mean prediction
time~$\overline{t}_{\text{pred.}}=\SI{1.124}{\micro\second}$. These correspond
to a standard error~$S=\num{0.013}$ and a relative speedup~$\omega=\num{6916416} \times$
with respect to the MC TBR evaluation baseline. Satisfying
goal~(b), the fastest model (Model~2, ANN) achieved~$R^2=\num{0.985}$,
$\overline{t}_{\text{pred.}}=\SI{0.898}{\micro\second}$, $S=\num{0.033}$
and~$\omega=\num{8659251} \times$.
While these surrogates
were trained on the entire available set of~\num{500000} datapoints, to satisfy
goal~(c) we also trained a more simplified model (Model~4, GBT)
that achieved~$R^2=\num{0.913}$,
$\overline{t}_{\text{pred.}}=\SI{6.125}{\micro\second}$, $S=\num{0.072}$ and $\omega=\num{1269777} \times$
with a set of size only~\num{10000}.

\begin{table*}
	\centering
	\sisetup{round-mode=places,round-precision=3,detect-weight=true,detect-family=true}
	\caption{\label{tbl:exp4-detailed-results}Results of experiment~4. Here,
		average values and standard deviations are reported over 5~cross-validation folds,
		$|\mathcal{T}|$~denotes cross-validation set size ($\times 10^3$)
		and $\omega$ is a relative speedup with respect to
		$\overline{t}_{\text{eval.}}=\SI{7.777049573054314}{\second}$
		measured in the MC TBR model during run~1 (see~\Tref{tbl:sampling-runs}).
	The best-performing method(s) under each are highlighted in bold.}
	\setlength\tabcolsep{4pt}
	\begin{indented}
	\item[]
		\scriptsize
		\begin{tabular}{lrrrrrrrr}
		\toprule
		{} & {} & \multicolumn{4}{c}{Regression performance} &
		\multicolumn{3}{c}{Complexity}\\
		\cmidrule(lr){3-6}
		\cmidrule(lr){7-9}
		Model & $|\mathcal{T}|$ & MAE [TBR] & $S$ [TBR] & $R^2$ [rel.] & $R^2_{\text{adj.}}$ [rel.]
						& $\overline{t}_{\text{trn.}}$ [\si{\milli\second}] &
		$\overline{t}_{\text{pred.}}$ [\si{\milli\second}] & $\omega$ [rel.]\\
		\midrule
		\input{tbl/table11}
		\bottomrule
		\end{tabular}
	\end{indented}
\end{table*}

@@

	\caption{Regression performance of Models 1-8 (from left to right, top to
		bottom) trained in the model comparison experiment, viewed
		as true vs.~predicted TBR on a test set of a selected cross-validation
		fold. Points are coloured by density.}
	\label{fig:reg-performance}
\end{figure}
